Supermartingales and Termination Analysis of Statistical PCF

### 2021-01-16 LICS21/SPCF.tex
Related work Introduction: ongoing Title Various sections: minor formatting so save space

The one-step reduction is given by the function $\red : \Lambda^0 \times \entrosp \to \Lambda^0 \times \entrosp$ where
\begin{equation*}
\red(M,s) := \left\{
    \begin{array}{ll}
        (E[N],s) & \text{if } M = E[R], R \to N\\
        & \qquad \text{ and } R \neq \tsample \\
        (E[\underline{\pi_h(s)}],\pi_t(s)) & \text{if } M = E[\tsample] \\
        (M,s) & \text{if } M \text{ is a value}
    \end{array} \right .
\end{equation*}
%\lo{Both $\to$ and $\red$ are called reduction relation.}

@@

The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid M \text{ terminates at } s \}$ is measurable. 
We say that a term $M$ is \emph{almost surely terminating} (AST) just if $\mu_{\entrosp}(\{s \mid M \text{ terminates at } s\}) = 1$.

\iffalse
\lo{Alternatively, define the \emph{runtime of $M$} to be the random variable 
\[
T_M(s) := 
\begin{cases}
\min \set{n \mid \pi_0(\red^n(M, s)) \textrm{ is a value}} & \hbox{if $\red^\infty(M,s)$ is defined}\\
\infty & \hbox{otherwise}
\end{cases}
\]
Equivalently, we say that $M$ is \emph{almost surely terminating} (AST) if $T_M < \infty$ a.s.; 
and $M$ is \emph{positively almost surely terminating} (PAST) if $\expect{T_M} < \infty$.}
\fi
For example, the term 
\[
(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)} ) \, \underline{0},
\] 
which generates a geometric distribution, terminates on the set $\entrosp \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely, whereas 
\[
\tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x. x) \underline 0}
\] 
which terminates on the set $\pi_h^{-1}[[0,0.5]]$, has probability 0.5 of failing to terminate.

@@

%\paragraph{}
\begin{remark}
\label{rem:score}
This definition of almost sure termination is equivalent to that given in \cite{MakOPW21} (although the program semantics is stated in a slightly different way), except for the presence of a $\tscore(r)$ construct for soft conditioning. 
The $\tscore(r)$ construct is not relevant to termination except that it fails if its argument is less than or equal to zero, which allows computations to fail after finitely many steps. 
If such a thing were added, it would merely require an additional side condition to our main theorems, that no failing term was reachable (or, in the case of the confluent semantics based results, no term where the reduction strategy is not defined is reachable via the reduction strategy unless it's a value).
\end{remark}

>>>>

\changed[lo]{The one-step reduction of the \emph{trace-based} (or \emph{sampling-style}) \emph{operational semantics} \cite{DBLP:journals/jcss/Kozen81,DBLP:conf/icfp/BorgstromLGS16,DBLP:conf/esop/CulpepperC17}} is given by the function $\red : \Lambda^0 \times \entrosp \to \Lambda^0 \times \entrosp$ where
\begin{equation*}
\red(M,s) := \left\{
    \begin{array}{ll}
        (E[N],s) & \text{if } M = E[R], R \to N\\
        & \qquad \text{ and } R \neq \tsample \\
        (E[\underline{\pi_h(s)}],\pi_t(s)) & \text{if } M = E[\tsample] \\
        (M,s) & \text{if } M \text{ is a value}
    \end{array} \right .
\end{equation*}
%\lo{Both $\to$ and $\red$ are called reduction relation.}

@@

The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid \hbox{$M$ terminates at $s$}\}$ is measurable. 
We say that a term $M$ is \emph{almost surely terminating} (AST) just if $\mu_{\entrosp}(\{s \mid \hbox{$M$ terminates at $s$}\}) = 1$.

\iffalse
\lo{Alternatively, define the \emph{runtime of $M$} to be the random variable 
\[
T_M(s) := 
\begin{cases}
\min \set{n \mid \pi_0(\red^n(M, s)) \textrm{ is a value}} & \hbox{if $\red^\infty(M,s)$ is defined}\\
\infty & \hbox{otherwise}
\end{cases}
\]
Equivalently, we say that $M$ is \emph{almost surely terminating} (AST) if $T_M < \infty$ a.s.; 
and $M$ is \emph{positively almost surely terminating} (PAST) if $\expect{T_M} < \infty$.}
\fi
For example, the term 
\[
(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)} ) \, \underline{0},
\] 
which generates a geometric distribution, terminates on the set $\entrosp \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely, whereas 
\(
\tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x. x) \underline 0}
\), 
which terminates on the set $\pi_h^{-1}[[0,0.5]]$, has probability 0.5 of failing to terminate.

@@

%\paragraph{}
\begin{remark}
\label{rem:score}
Our definition of AST is equivalent to that given in \cite{MakOPW21} (although the program semantics is stated in a slightly different way), except for the presence of a $\tscore(r)$ construct for soft conditioning. 
The $\tscore(r)$ construct is irrelevant to termination except that it fails if its argument is less than or equal to zero, which allows computations to fail after finitely many steps. 
If such a thing were added, it would merely require an additional side condition to our main theorems, that no failing term was reachable (or, in the case of the confluent semantics based results, no term where the reduction strategy is not defined is reachable via the reduction strategy unless it's a value).
\end{remark}

### 2021-01-16 LICS21/antitone.tex
Related work Introduction: ongoing Title Various sections: minor formatting so save space

\paragraph*{Completeness}
It seems very likely that this method is complete in the case that every reachable term is AST, but we have been unable to actually prove this. It is certainly at least capable of proving the termination of terms which terminate arbitrarily slowly (in the sense that there is no similar limitation to Theorem \ref{thm:rankable implies termination}, which can only prove termination of $\tY$-PAST terms).

>>>>

\paragraph*{Completeness}
It seems very likely that this method is complete in the case that every reachable term is AST (\cref{conj:antitone}), but we have been unable to actually prove this. It is certainly at least capable of proving the termination of terms which terminate arbitrarily slowly (in the sense that there is no similar limitation to Theorem \ref{thm:rankable implies termination}, which can only prove termination of $\tY$-PAST terms).

### 2021-01-16 LICS21/confluent.tex
Related work Introduction: ongoing Title Various sections: minor formatting so save space

When proving almost sure termination in this way, it is necessary to consider all of the terms a given term may reduce to, and organise them into a manageable number of cases to assign ranking function values. Sometimes however, the reduction that the programmer has in mind may not be strictly the call-by-value order defined so far; or the number of cases necessary for defining a sparse ranking function is excessive because the reduction does not proceed in a neat and orderly manner. In cases such as these, it may be much more convenient to instead consider ranking functions with respect to alternative reduction strategies constructed to match the individual term in question, thereby reducing the number of cases needed and simplifying the reachable terms. It will be shown in \Cref{thm:confluent ranking} that ranking functions with respect to alternative reduction strategies are sound for proving AST in the usual sense (i.e.~with respect to the standard CBV reduction strategy). In order to obtain this result, we prove a relation between the results of using different reduction strategies, by first constructing a confluent trace semantics that is equivalent to the standard semantics as a special case.

@@

A \emph{position} is a finite sequence of steps into a term, defined inductively as
\begin{align*}
\alpha ::= & \; {\cdot} \mid \lambda ; \alpha \mid @_1 ; \alpha \mid @_2 ; \alpha \mid \underline f_i ; \alpha \\
& {} \mid \tY ; \alpha \mid \textsf{if}_1 ; \alpha \mid \textsf{if}_2 ; \alpha \mid \textsf{if}_3 ; \alpha.
\end{align*}
The \emph{subterm of $M$ at $\alpha$}, denoted $M \mid \alpha$, is defined as
\begin{align*}
M \mid \cdot & = M \\
\lambda x. M \mid \lambda ; \alpha & = M \mid \alpha \\
M_1 M_2 \mid @_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2 \\
\underline f(M_1,\dots,M_n) \mid \underline f_i ; \alpha & = M_i \mid \alpha \quad \text{for }i \leq n \\
\tY M \mid \tY ; \alpha & = M \mid \alpha \\
\tif{M_1 < 0}{M_2}{M_3} \mid \textsf{if}_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2,3
\end{align*}
so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g. $x \, y \mid \lambda$ is undefined). 
A position such that $M\mid \alpha$ does exist is said to \emph{occur} in $M$. 
%\changed[lo]{
\emph{Substitution} of $N$ at position $\alpha$ in $M$, written $M[N/\alpha]$, is defined similarly.
For example, let 
\[
M = \lambda x \, y. y \, (\tif{x < 0}{y \, (\underline f (x))}{\underline 3})
\quad
\alpha =\lambda ; \lambda ; @_2 ; \textsf{if}_2 ; @_2
\]
then 
\(
M[\tsample / \alpha] = \lambda x \, y. y \, (\tif{x < 0}{y \, \tsample}{\underline 3}).
\)
%}

>>>>

When proving almost sure termination in this way, it is necessary to consider all of the terms a given term may reduce to, and organise them into a manageable number of cases to assign ranking function values. 
Sometimes however, the reduction that the programmer has in mind may not be strictly the call-by-value order defined so far; 
or the number of cases necessary for defining a sparse ranking function is excessive because the reduction does not proceed in a neat and orderly manner. 
In cases such as these (e.g.~\cref{ex:sum of powers}), it may be much more convenient to instead consider ranking functions with respect to alternative reduction strategies constructed to match the individual term in question, 
thereby reducing the number of cases needed and simplifying the reachable terms. 
It will be shown in \Cref{thm:confluent ranking} that ranking functions with respect to alternative reduction strategies are sound for proving AST in the usual sense (i.e.~with respect to the standard CBV reduction strategy). 
In order to obtain this result, we prove a relation between the results of using different reduction strategies, 
by first constructing a confluent trace semantics that is equivalent to the standard semantics as a special case.

@@

A \emph{position} is a finite sequence of steps into a term, defined inductively as
\begin{align*}
\alpha ::= & \; {\cdot} \mid \lambda ; \alpha \mid @_1 ; \alpha \mid @_2 ; \alpha \mid \underline f_i ; \alpha \\
& {} \mid \tY ; \alpha \mid \textsf{if}_1 ; \alpha \mid \textsf{if}_2 ; \alpha \mid \textsf{if}_3 ; \alpha.
\end{align*}
The \emph{subterm of $M$ at $\alpha$}, denoted $M \mid \alpha$, is defined as
\begin{align*}
M \mid \cdot & = M \\
\lambda x. M \mid \lambda ; \alpha & = M \mid \alpha \\
M_1 M_2 \mid @_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2 \\
\underline f(M_1,\dots,M_n) \mid \underline f_i ; \alpha & = M_i \mid \alpha \quad \text{for }i \leq n \\
\tY M \mid \tY ; \alpha & = M \mid \alpha \\
\tif{M_1 < 0}{M_2}{M_3} \mid \textsf{if}_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2,3
\end{align*}
so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g. $x \, y \mid \lambda$ is undefined). 
A position such that $M\mid \alpha$ does exist is said to \emph{occur} in $M$. 
%\changed[lo]{
\emph{Substitution} of $N$ at position $\alpha$ in $M$, written $M[N/\alpha]$, is defined similarly.
For example, let 
\(
M = \lambda x \, y. y \, (\tif{x < 0}{y \, (\underline f (x))}{\underline 3})
\)
and
\(
\alpha =\lambda ; \lambda ; @_2 ; \textsf{if}_2 ; @_2
\)
then 
\(
M[\tsample / \alpha] = \lambda x \, y. y \, (\tif{x < 0}{y \, \tsample}{\underline 3}).
\)
%}

### 2021-01-16 LICS21/intro.tex
Related work Introduction: ongoing Title Various sections: minor formatting so save space

Various theorems about probabilistic programs rely on the assumption that the program terminates almost surely. One possible approach to prove almost sure termination (AST) is to find some variant on the program state that decreases on average sufficiently quickly that it must at some point reach 0, at which point the program terminates. In other words, the program's behavior is used to define an associated (ranking) supermartingale. Proof rules based on relating the program state to supermartingales already exist for first-order imperative programs \cite{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18}. This paper's contribution is to extend this method to a higher-order setting.

The language PPCF used in this paper is simply typed, with an explicit recursion primitive, $\tY$. As the type system already constrains the terms enough to force termination in the absence of the recursion construct \cite{tait1967,BarendregtDS2010}, it is only the $\tY$-reduction steps that must be counted, making defining ranking functions somewhat easier.

The notion of a \emph{sparse ranking function} is introduced to make constructing ranking functions more convenient. Most of the individual execution steps of a typical program are trivial and easy to mentally skip over. Sparse ranking functions can be defined only for those points in the execution of a program which are semantically important, while all of the other intermediate steps can be ignored. This also makes the ranking function method of proving AST more compatible with syntactic sugar, because the intermediate reduction steps implicit in the simplified notation can be ignored.

A ranking function (or sparse ranking function) provides a bound on the expected number of $\tY$-reduction steps before the program terminates, therefore ranking functions cannot be constructed for terms whose expected number of $\tY$-reduction steps is infinite, such as the simplest implementation of the 1D unbiased random walk. This restriction can be removed by generalising ranking functions to antitone ranking functions, which rather than having to decrease by a constant amount for each $\tY$-reduction step, may decrease by a variable amount, depending on the value of the ranking function. This method is capable of proving the almost sure termination of programs which terminate arbitrarily slowly. Our progress condition is equivalent to that given in \cite{DBLP:journals/pacmpl/McIverMKK18} (by applying a transformation to the ranking function), but needs to be different in order to be compatible with the presence of additional intermediate reduction steps where the ranking function is not required to decrease at all.

Basic (non-random) lambda calculus has the very useful Church-Rosser property, which implies (among other things) that even if execution of a program starts in a different order, it will still reach the same normal form eventually (assuming it does reach a normal form). Probabilistic lambda calculus does not have this property, because random choices may be duplicated, and evaluating the same sub-term multiple times can yield different results. However, with a restricted set of reduction strategies, Church-Rosserness may be regained. A novel addressing scheme for the possible random choices in a program's execution is introduced, which ensures that the same random choices are taken at corresponding positions in alternative reduction sequences, so that the same eventual result can be reached. This is then used to prove yet another extension to the ranking function theorem, that ranking functions may be defined with respect to alternative reduction strategies (which in some cases may lead to a considerably simpler execution and ranking function), and rankability in this sense still imples almost sure termination. This confluent trace semantics has other possible applications as well, for example in Bayesian inference algorithms.

@@

We present the syntax and trace (or sampling) semantics of PPCF in \Cref{sec:PPCF}.
In \Cref{sec:supermartingales}, we show that ranking functions on terms induce supermartingales, which form the basis of a sound method for proving AST.
We introduce sparse ranking functions in \Cref{sec:ranking} and antitone ranking functions in \Cref{sec:antitone}, and illustrate how they can be used to prove AST via examples. 
In \Cref{sec:confluent}, we present a confluent trace semantics and demonstrate its usefulness.
We discuss further applications of the confluent semantics in \Cref{sec:applications}; and conclude with comments on related work and further directions in \Cref{sec:related}.

>>>>

Various theorems about probabilistic programs rely on the assumption that the program terminates almost surely. One possible approach to prove almost sure termination (AST) is to find some variant on the program state, called \emph{ranking function}, that decreases on average sufficiently quickly that it must at some point reach 0, at which point the program terminates. In other words, the program's behavior is used to define an associated (ranking) \emph{supermartingale}. Proof rules based on relating the program state to supermartingales already exist for first-order imperative programs \cite{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18}. This paper's contribution is to extend this method to a higher-order setting.

The language PPCF used in this paper is simply typed, with sampling from a continuous distribution, and an explicit recursion primitive, $\tY$. 
As the type system already constrains the terms enough to force termination in the absence of the recursion construct \cite{tait1967,BarendregtDS2010}, it is only the $\tY$-reduction steps that must be counted, making defining ranking functions somewhat easier.

We introduce the notion of \emph{sparse ranking function} to make constructing ranking functions more convenient. Most of the individual execution steps of a typical program are trivial and easy to mentally skip over. Sparse ranking functions can be defined only for those points in the execution of a program which are semantically important, while all of the other intermediate steps can be ignored. This also makes the ranking function method of proving AST more compatible with syntactic sugar, because the intermediate reduction steps implicit in the simplified notation can be ignored.

A ranking function (or sparse ranking function) provides a bound on the expected number of $\tY$-reduction steps before the program terminates, therefore ranking functions cannot be constructed for terms whose expected number of $\tY$-reduction steps is infinite, such as the simplest implementation of the 1D unbiased random walk. 
This restriction can be removed by generalising ranking functions to \emph{antitone ranking functions}, which rather than having to decrease by a constant amount for each $\tY$-reduction step, may decrease by a variable amount, depending on the value of the ranking function. 
This method is capable of proving the almost sure termination of programs which terminate arbitrarily slowly. Our progress condition is equivalent to that given by McIver et al.~\cite{DBLP:journals/pacmpl/McIverMKK18} (by applying a transformation to the ranking function), but needs to be different in order to be compatible with the presence of additional intermediate reduction steps where the ranking function is not required to decrease at all.

Basic (non-random) lambda calculus has the very useful Church-Rosser property, which implies (among other things) that even if execution of a program starts in a different order, it will still reach the same normal form eventually (assuming it does reach a normal form). 
Probabilistic lambda calculus does not have this property, because random choices may be duplicated, and evaluating the same subterm multiple times can yield different results. 
However, with a restricted set of reduction strategies, Church-Rosserness may be regained. 
We introduce a novel addressing scheme for the possible random choices in a program's execution, which ensures that the same random choices are taken at corresponding positions in alternative reduction sequences, so that the same eventual result can be reached. 
This is then used to prove yet another extension to the ranking function theorem, that ranking functions may be defined with respect to alternative reduction strategies (which in some cases may lead to a considerably simpler execution and ranking function), and rankability in this sense still imples almost sure termination. 
This confluent trace semantics has other possible applications as well, for example in Bayesian inference algorithms.

@@

We present the syntax and trace semantics of PPCF in \Cref{sec:PPCF}.
In \Cref{sec:supermartingales}, we show that ranking functions on terms induce supermartingales, which form the basis of a sound method for proving AST.
We introduce sparse ranking functions in \Cref{sec:ranking} and antitone ranking functions in \Cref{sec:antitone}, and illustrate how they can be used to prove AST via examples. 
In \Cref{sec:confluent}, we present a confluent trace semantics and demonstrate its usefulness.
We discuss further applications of the confluent semantics in \Cref{sec:applications}; and conclude with comments on related work and further directions in \Cref{sec:related}.

### 2021-01-16 LICS21/ranking.tex
Related work Introduction: ongoing Title Various sections: minor formatting so save space

Although rankability implies almost sure termination, the converse does not hold in general. For example,
\begin{equation}
\tif{{-}\tsample < 0}{\underline{0}}{(\tY \lambda x. x) \, \underline 0}
\label{ex:0 probability reachable}
\end{equation}
terminates in 3 steps with probability 1, but isn't rankable because $(\tY \lambda x. x) \, \underline 0$ is reachable, although that has probability 0. 
Not only is this counterexample AST, it's PAST. %positively almost surely terminating i.e.~the expected time to termination is finite.

A ranking function can be constructed under the stronger assumptions that, for every $N$ reachable from $M$, the expected number of $\tY$-reduction steps from $N$ to a value is finite. 
In particular, the expected number of $\tY$-reduction steps from each reachable term is a ranking function. 
However, note that a finite number of expected $\tY$-reduction steps does not necessarily imply a finite number of expected total reduction steps.

>>>>

Although rankability implies almost sure termination, the converse does not hold in general. For example,
\begin{equation*}
\tif{{-}\tsample < 0}{\underline{0}}{(\tY \lambda x. x) \, \underline 0}
\label{ex:0 probability reachable}
\end{equation*}
terminates in 3 steps with probability 1, but isn't rankable because $(\tY \lambda x. x) \, \underline 0$ is reachable, although that has probability 0. 
Not only is this counterexample AST, it's PAST. %positively almost surely terminating i.e.~the expected time to termination is finite.

A ranking function can be constructed under the stronger assumptions that, for every $N$ reachable from $M$, the expected number of $\tY$-reduction steps from $N$ to a value is finite. 
In particular, the expected number of $\tY$-reduction steps from each reachable term is a ranking function. 
However, a finite number of expected $\tY$-reduction steps does not necessarily imply a finite number of expected total reduction steps.

### 2021-01-16 LICS21/related.tex
Related work Introduction: ongoing Title Various sections: minor formatting so save space

Most methods for proving AST of programs do not support continuous distributions \cite{DBLP:journals/toplas/LagoG19,DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/lics/KobayashiLG19,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05}.
The few that do \cite{DBLP:conf/popl/FioritiH15,DBLP:conf/pldi/ChenH20,DBLP:journals/toplas/ChatterjeeFNH18} are for first-order imperative programs.

A fortiori, our results specialise to a method for proving termination of non-random lambda calculus. 
Thus restricted, our approach is closest in spirit to Jones' work on size-change termination \cite{DBLP:journals/lmcs/JonesB08,DBLP:conf/aplas/SereniJ05}. 
Dal Lago and Grellois \cite{DBLP:journals/toplas/LagoG19} extend this to programs with discrete distributions, but their method is restricted to first-order affine recursion.
Breuvart and Dal Lago \cite{DBLP:conf/ppdp/BreuvartL18} develop systems of intersection types from which the termination probability of higher-order programs (with discrete distributions) can be inferred from (infinitely many) type derivations.

Kobayashi et al.~\cite{DBLP:conf/lics/KobayashiLG19} show that the termination probability of CbN order-$n$ probabilistic recursion schemes ($n$-PHORS) can be obtained as the least fixpoint of suitable order-$(n-1)$ fixpoint equations, which are solvable using Kleene fixpoint iteration.  
By contrast, our approach works for programs with continuous distributions. 
It is worth noting that (order-$n$) PHORS is strictly less expressive then (order-$n$) CbN PPCF (because the underlying recursion schemes are not Turing complete). 
Some interesting PPCF terms such as \cref{ex:raven complex,ex:higher-order recursion}
%\refExample{complexExample} 
are not expressible as PHORS.

The main result in the work \cite{DBLP:journals/pacmpl/McIverMKK18} by McIver et al.~is similar to our \cref{thm:antitone rankable implies termination}, but in an imperative language. While we require that the antitone ranking function decreases for every $\tY$-reduction step, they require that it decreases for every iteration of a certain while loop (the ranking function in that case is defined in the context of a particular loop), which is similar to $\tY$ but limited to tail recursion. 
The difference in the exact progress condition is not significant. 
\akr{Their progress condition, that there are antitone functions $p$ and $d$ such that the ranking function decreases from $x$ by at least $d(x)$ with probability at least $p(x)$, would not have worked directly because it does not cope as easily with the fact that the ranking function may change in value at the intermediate non-$\tY$ reduction steps (provided it does not increase in expectation). However, it is easy enough to convert a ranking function that satisfies either progress condition to the other.}
\lo{We may have to omit this.}

@@

\paragraph*{Further directions} We have not yet been able to prove \Cref{conj:antitone} which would imply that \Cref{thm:antitone rankable implies termination} is almost complete.
\begin{conjecture}
\label{conj:antitone}
Every term is antitone rankable if all of the terms reachable from it are AST.
\end{conjecture}

>>>>

Most methods for proving AST of programs in the literature do not support continuous distributions \cite{DBLP:journals/toplas/LagoG19,DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/lics/KobayashiLG19,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05}.
The few that do \cite{DBLP:conf/popl/FioritiH15,DBLP:conf/pldi/ChenH20,DBLP:journals/toplas/ChatterjeeFNH18} are for first-order imperative programs.

A fortiori, our results specialise to a method for proving termination of non-random lambda calculus. 
Thus restricted, our approach is closest in spirit to Jones' work on size-change termination \cite{DBLP:journals/lmcs/JonesB08,DBLP:conf/aplas/SereniJ05}. 
Dal Lago and Grellois \cite{DBLP:journals/toplas/LagoG19} extend this to programs with discrete distributions, but their method is limited to affine recursion of first-order functions.
Breuvart and Dal Lago \cite{DBLP:conf/ppdp/BreuvartL18} develop systems of intersection types from which the termination probability of higher-order programs (with discrete distributions) can be inferred from (infinitely many) type derivations.

Kobayashi et al.~\cite{DBLP:conf/lics/KobayashiLG19} show that the termination probability of order-$n$ probabilistic recursion schemes ($n$-PHORS) can be obtained as the least fixpoint of suitable order-$(n-1)$ fixpoint equations, which are solvable by Kleene fixpoint iteration.  
By contrast, our approach works for programs with continuous distributions. 
Note that $n$-PHORS is strictly less expressive then order-$n$ CbN PPCF (because the underlying recursion schemes are not Turing complete):
PPCF terms in \cref{ex:raven complex,ex:higher-order recursion}
%\refExample{complexExample} 
are not definable as PHORS.

The main result by McIver et al.~\cite{DBLP:journals/pacmpl/McIverMKK18} is similar to our \cref{thm:antitone rankable implies termination}, but in an imperative language. While we require that the antitone ranking function decreases for every $\tY$-reduction step, they require that it decreases for every iteration of a certain while loop (the ranking function in that case is defined in the context of a particular loop), which is similar to $\tY$ but limited to tail recursion. 
The difference in the exact progress condition is not significant. 
\akr{Their progress condition, that there are antitone functions $p$ and $d$ such that the ranking function decreases from $x$ by at least $d(x)$ with probability at least $p(x)$, would not have worked directly because it does not cope as easily with the fact that the ranking function may change in value at the intermediate non-$\tY$ reduction steps (provided it does not increase in expectation). However, it is easy enough to convert a ranking function that satisfies either progress condition to the other.}
\lo{Can you replace the highlighted by a brief comment on their discussion of completeness?}

@@

\paragraph*{Further directions} We have not yet been able to prove \Cref{conj:antitone} which would imply that \Cref{thm:antitone rankable implies termination} is almost complete.
\begin{conjecture}[Completeness]
\label{conj:antitone}
Every PPCF term is antitone rankable if all terms reachable from it are AST.
\end{conjecture}
